---
title: "02-tidy"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r}
library(readr)
library(tidyr)
library(dplyr)
library(here)
```

```{r}
pew <- readr::read_csv(here::here("data", "pew.csv"))
pew
```

```{r}
head(pew)
```

## Column headers are values, not variable names.

```{r}
pew %>%
  tidyr::pivot_longer(cols = -religion)
```

```{r}
pew_tidy <- pew %>%
  tidyr::pivot_longer(cols = -religion, names_to = "income", values_to = "count")
```


```{r}
billboard <- readr::read_csv(here::here("data", "billboard.csv"))
```

```{r}
head(billboard)
```

```{r}
billboard_tidy <- billboard %>%
  tidyr::pivot_longer(cols = wk1:wk76,
                      names_to = "week",
                      values_to = "rank")
```

```{r}
billboard_tidy
```


```{r}
billboard_tidy <- billboard %>%
  tidyr::pivot_longer(cols = wk1:wk76,
                      names_to = "week",
                      values_to = "rank",
                      names_prefix = "wk",
                      values_drop_na = TRUE
                      )
billboard_tidy
```

```{r}
billboard_tidy %>%
  dplyr::group_by(week) %>%
  summarize(avg_rnk = mean(rank))
```

```{r}
billboard_tidy %>%
  dplyr::mutate(week = as.numeric(week)) %>%
  dplyr::group_by(week) %>%
  summarize(avg_rnk = mean(rank))
```


## Multiple variables are stored in one column

```{r}
ebola <- readr::read_csv(here::here("data", "country_timeseries.csv"))
ebola
```

```{r}
ebola_tidy <-ebola %>%
  tidyr::pivot_longer(
    cols = tidyselect::contains("_"),
    names_to = c("type", "country"),
    names_pattern = "(.*)_(.*)",
    values_to = "count"
)
```

```{r}
ebola_tidy
```


```{r}
ebola_tidy <-ebola %>%
  tidyr::pivot_longer(
    cols = !Date:Day,
    names_to = c("type", "country"),
    names_pattern = "(.*)_(.*)",
    values_to = "count"
)
ebola_tidy
```

```{r}
ebola_tidy <-ebola %>%
  tidyr::pivot_longer(
    cols = Cases_Guinea:tidyselect::last_col(),
    names_to = c("type", "country"),
    names_pattern = "(.*)_(.*)",
    values_to = "count"
)
ebola_tidy
```

## Variables are stored in both rows and columns.


```{r}
weather <- readr::read_csv(here::here("data", "weather.csv"))
weather
```

```{r}
weather_long <- weather %>%
  tidyr::pivot_longer(
    cols = tidyselect::starts_with("d"),
    names_to = "day",
    names_prefix = "d",
    values_to = "temperature",
    values_drop_na = TRUE
  )
weather_long
```

```{r}
weather_tidy <- weather_long %>%
  tidyr::pivot_wider(names_from = "element", values_from = "temperature")
```

## Exercises

1. Tidy the `who` datadset from the `tidyr` package by making it longer and breaking up the columns from `new_sp_m014` where

- "diagnosis" = `sp`
- "gender" = m
- "age" = 014

```{r}
who
```


```{r}
who %>% pivot_longer(
  cols = _____,
  names_to = c(_____),
  names_pattern = "new_?(.*)_(.)(.*)",
  values_to = _____,
  values_drop_na = TRUE
)
```

2. pivot wider the `us_rent_income`, hint: you can pass in multiple values for `values_from`

```{r}
us_rent_income
```


```{r}
us_rent_income %>%
  pivot_wider(names_from = _____, values_from = c(_____, _____))
```

